---
title: "DiseaseDiva"
author: "Us"
date: "10/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 
Prostate cancer is among one of the leading causes of death in males in the United States. The genomic study of metastatic prostate cancer has been limited due to tumor samples being obtained from autopsies as well as limited cohort preclinical models. Prospective studies are few and far between due to limitations in collecting adequate tumor tissue through biopsy. A paper published in 2015 prospectively analyzed tumor biopsies among affected individuals. Subsequently as of 2016, genomic data for prostate cancer has become more widely available due to the National Cancer Institutes (NCI) Genomics Data Commons (GDC).

In this project we will explore data from both the 2015 prospective cohort study along with data available from the NCI's Genomic Data Commons. This project aims to investigate three datasets, through mapping and transformation of the raw data in multiple programs such as Excel, SQL and R,  with the ultimate goal of optimizing downstream analysis of the prostate cancer cohorts. We will discuss various methods by which we wrangled the data to investigate underlying associations between individual case presentations and manifestation of disease.

A team of 73 researchers conducted a multi-institutional study that investigated clinical genomics of advanced prostate cancer specifically metastatic, castration resistant prostate cancer (mCRPC). mCRPC occurs in individuals who develop resistance to androgen deprivation therapies (ADT). The aim of this research was to facilitate precision medicine by developing a prospective whole-exome and transcriptome sequencing of tumor biopsies. Data was obtained from living affected individuals to compile a set of genomic alterations. The study was published in the journal "Cell" in May 2015. 

The link to the publication can be found [here.](https://pubmed.ncbi.nlm.nih.gov/26000489/)

Data was extracted from the cBioPortal, a public cancer genomics site hosted and maintained by the Memorial Sloan Kettering Cancer Center. The sample included 150 individuals with metastatic prostate cancer along with 17 data points per individual,  described below:
-**Patient ID**: unique patient identifier (number)
-**Sample ID**: unique biopsy sample identifier from clinical trial(number)
-**Mutation Count**: mutation rate for mCRPC defined as mutations per megabase or mutations/Mb (number)
-**Fraction Genome Altered**: fraction of mutant allele variants in biopsy (number)
-**Diagnosis Age**: patient's age when they were diagnosed (number)
-**Tumor Site**: physical site of tumor on human body from which biopsy was taken (string)
-**Abiraterzone (ABI) and Enzalutamide (ENZA) Exposure Status**: exposure to  second-generation androgen deprivation therapies (binary: yes/no)
-**Center of Sequencing**: the international academic medical center where exome sequencing occurred (string)
-**Prior Treatment**: receipt of previous androgen deprivation therapy (binary:yes/no)
-**Site Sequenced**: site of exome sequencing: Broad or UM (string)
-**tumor content**: percentage of tumor captured in the biopsy (number)

The National Cancer Institute (NCI) maintains a unified repository of genomic data known as the Genomic Data Commons(GDC). Groundwork for this repository began in June 2015 with nearly 50,000 raw data sequences analyzed by June 2016. This data sharing platform uses an Open-Stack-based private cloud to unite several large-scale cancer genome research programs such as TCGA and OCG. The purpose of this repository is to standardize the data processing of these multi-dimensional datasets by using common bioinformatics pipelines that facilitate direct comparison of the data. 

More about the National Cancer Institute's Genomic Data Commons can be found [here.](https://gdc.cancer.gov/about-gdc)

Data was extracted from GDC Portal's Exploration page. The "cleaned" version of the data sample included 493 individuals with metastatic prostrate cancer along with 24 data points per individual as described below: 
-**CaseUUID**: universally unique identifier assigned to every entity in the data model (string)
-**CaseID**:specific individual cases using the submitter ID from the genomic cancer research program (string)
-**Project**:The cancer research program from which the case originated from (string)
-**Primary Site**: The primary site of the cancer (string)
-**Gender**:"Gender is defined as a set of characteristics that distinguish persons based on their social roles" (string)
-**Files**: the quantity of files available for that case (number)
-**Seq**:Number of the sequencing reads(an inferred sequence of basepairs) (number)
-**Exp**:The transcriptome profiling (number)
-**SNV**:simple nucleotide variation (number)
-**CNV**:Copy number variation,the number of copies of a particular gene varies from one individual to the next (number)
-**Meth**:Number of DNA methylation that took place (number)
-**Clinical**: The number of clinical alterations (number)
-**Bio**:biospecimen (number)
-**Mutations**:The number of simple somatic mutations detected for that case(number)
-**Genes**: The quantity of genes affected by mutations for each case (number)
-**Slides**: the quantity of slides available for each case (number)
-**Program**: The type of program the subject is enrolled in “The Cancer Genome Atlas” or the “Count Me In” (string) 
-**Disease Type**: This describes if it is Adenomas and Adenocarcinomas, the location where the cancer is formed (string)
-**Age at diagnosis**:Age at the time of diagnosis calculated from their day of
birth to the day they were diagnosed with cancer (string)
-**Days to death**:The date from when they were diagnosed, and they died (string)
-**Vital Status**: the survival status of the person (string)
-**Primary Diagnosis**:Their initial pathological diagnoses of cancer (string)
-**Ethnicity**: "An individual's self-identified social and cultural category, particularly whether they identify as Hispanic or Latino" (string)
-**Race**: "A taxonomic group that is a division of a species that is classified arbitrarily. It is characterized by shared heredity, physical attributes, and behavior, and in the case of humans, by common history, nationality, or geographic distribution" (string)


##Methods 

##Extract the Data from the cBioPortal 
The first dataset was downloaded from the cBioPortal site using the "download clinical data for the selected cases" button extracting all columns. 

The link to the dataset can be found [here.](https://www.cbioportal.org/study/clinicalData?id=prad_su2c_2015)

##Cleaning Our Data
The first dataset was downloaded as a TSV file which required conversion in Excel to a CSV file. After importing the data into Excel using the Import Data From Text tool, the worksheet was saved as a csv file. The raw data contained 150 rows and 20 columns. As a team we worked through the exploratory investigation and cleaning of the data by utilizing GitHub to collaborate on an R Markdown file. 

```{r}
#Loaded the data as a CSV into the R Markdown file 
rawdat<- read.csv("~/Documents/Dartmouth/Fall 2021/QBS 181/group project/Data .csv")
```

To gain an understanding for the types of variables we would be analyzing along with some of the numeric summaries for the continuous variables, we employed the summary function. 

```{r pressure, echo=FALSE}
#Summarize the variables 
summary(rawdat)
```
Preliminary findings indicate, mean age of diagnosis for prostate cancer is approximately 67 years of age. Average tumor content found in these biopsies appears to be 60%. Initial analysis revealed that there were several cells for which NA values were present. The following command was used to eliminate the NAs saving the new data to a dataframe called complete_dat. 

```{r}
#Remove NAs
complete_dat <- na.omit(rawdat)
```

Upon removal of the NAs, the dataset reduced to 147 rows and 20 columns. Three patients from the original extract contained in rows 71 to 73 were therefore excluded from subsequent analyses due to the NAs being present in both the Diagnosis.Age column along with the Abiraterone (ABI) and Enzalutamide (ENZA) Exposure Status. These two columns are crucial in further analyses being as age is a related factor for severity of disease. Furthermore, without information as to prior exposure status to the androgen deprivation therapies, assessment of mCRPC is incomplete. 

Further analysis of the data revealed that several columns were found to have duplicated values. To assess whether the individual cases included in the sample were duplicated, we utilized the duplicated command to check the Sample ID column for repeat values. As shown below, none were found. 

```{r}
#Check for duplicates
duplicated(complete_dat$Sample.ID)
```
Upon cursory glance the Patient ID and Sample ID column appeared to be populated with the exact same values per each row. To confirm this assumption we used the following syntax to check if the values between the two columns were unique. The syntax reads as follows, from the compelete_dat dataframe check if Patient Id is not contained in Sample Id then output the results. 

```{r}
#Check if sample id is unique
complete_dat[!complete_dat$Patient.ID %in% complete_dat$Sample.ID, ]
```

The output above is empty indicating that all the Patient Ids are contained within the Sample Id. In other words the two columns are the same. 

Several columns appeared to be populated by the same value for each row. For example, the Cancer.Type column was entirely populated by the value "Prostate Adenocarcinoma".  To determine whether this assumption was indeed true we created the following function utilizing a for loop to cycle through each column to check that the values contained in the column were unique. 

```{r}
find_nonuniq <- function(dataset){ #defining the function 
  cols_nonunique <- c() #set an empty vector to contain the results of the function
  for(i in 1:ncol(dataset)){ #initiated the for loop to traverse all the columns
    if(length(unique(dataset[,i])) <= 1){ #if the output of the unique function showed 1 or fewer values add that column index to the vector 
      cols_nonunique[length(cols_nonunique) + 1] <- i #append the vector with the new columns 
    }
  }
  print(cols_nonunique) #print the vector 
} 
cols_nonunique <- find_nonuniq(complete_dat) #run the function on the complete_dat dataframe 

```
After, running this function the output indicated that 9 columns were found to contain the same repeated value for every row contained within that column. To verify that this was true we cross checked each column index against the "complete_dat" dataframe and found that indeed the values were repeated by row. We therefore removed each of those columns according to the index presented in the cols_nonunique output. 

```{r}
remove_nonuniques <- complete_dat[,-cols_nonunique] #removing the columns with repeated values 
```

Consquently, after removal of the data columns with repeated values we were left with the "remove_nonuniques" dataframe which contained 147 rows and 11 columns. 

##Bonus Aim 
The last aim checks for an association between diagnosis age and tumor site. According to the Cleveland Clinic 1 in 6 men will be diagnosed with prostate cancer during the course of their lifetime. Research from the American Cancer Society suggests that 6 out of 10 cases of prostate cancer are diagnosed among men 65 years or older. As mentioned previously the mean age of diagnosis among this prospective cohort was 67 years of age. Prior research suggests that as men get older their risk of being diagnosed with prostate cancer increases. In addition, the publication in Cell suggests that bone tumors are the most common site of metastatic disease. This aim therefore intends to determine whether these two prior notions hold true among the data in question. 

To conduct the analysis we subset the data columns Diagnosis.Age and Tumor.Site into a dataframe called age_tumor. The numeric index of these columns was 4 and 11 respectively. Before proceeding with any analysis we conducted an exploratory investigation to understand the quantity of unique tumor sites present in the data set along with the spread of age at diagnosis.  
```{R}
age_tumor <- remove_nonuniques[c(4,11)] #subset the data 
age_tumor
summary(unique(age_tumor$Tumor.Site)) #look at unique values for tumor.site 
```

According to the aforementioned summary and unique command, 20 unique tumor sites are present within this data. Next, we explored the distribution of diagnosis age to determine min, max, mean and median ages for this prospective prostate cohort. 

```{r}
summary(age_tumor$Diagnosis.Age)
```
Given the minimum age at diagnosis was 40 and the maximum age at diagnosis was 85 we needed to come up with a way to age-stratify the data in order to compare the various tumor sites with age at diagnosis. TAn article published in the [British Journal of Cancer](https://www.nature.com/articles/bjc2013268#Tab1) described a methodology for stratifying age within a population based study investigating prostate cancer. The researchers stratified age according to the following groups "40-59", "60-69", "70-79" and "80+ years". Since this was a validated study we employed the same age-stratified bins for our analysis. 

```{r}
library(tidyverse) #load tidyverse to assist in analysis 
#grouping the age by the BJC age-stratified bins using ifelse statements 
age_tumor$agegrp <- ifelse(age_tumor$Diagnosis.Age >= 40 & age_tumor$Diagnosis.Age <=59, "40-59",
                  ifelse(age_tumor$Diagnosis.Age > 59 & age_tumor$Diagnosis.Age <= 69, "60-69",
                         ifelse(age_tumor$Diagnosis.Age > 69 & age_tumor$Diagnosis.Age <= 79, "70-79",
                                ifelse(age_tumor$Diagnosis.Age >79,"80+", NA))))
```

After stratifying the data, we created a summary table to assess the aggregated counts of tumors by site grouped by age. 

```{r}
library(dplyr) #load dplyr to use gtsummary 
#install.packages("gtsummary") #in order to create the graph you may need to install the gtsummary pkg
library(gtsummary) #load gtsummary to create the summary table 
age_tumor %>% 
  select(Tumor.Site, agegrp) %>%
  tbl_summary(by = agegrp, label = Tumor.Site ~ "Tumor Site") %>% #develop the table based on the selected input
  add_p(test=everything()~"kruskal.test") #test for the differences among ranks 
```
As evidenced by the output of the table, several of the aggregate counts of tumors by site were either 0 or 1 in most cases, which is not that significant. Five sites however did demonstrate counts greater than 1 and those include, Bone, Liver, Lymph Node, Prostate and Soft Tissue. To assess whether the differences among the age stratified ranks were significant by tumor site we conducted a Kruskal-Wallis Rank Sum Test. The output of which can be found in the last column of this table. The value of p was reported as 0.9 and assuming an alpha significance level of 0.05 this value is not significant. Thus, we conclude that there is no significant difference exhibited among the age stratified ranks by tumor site. 

We found the quantities of the aggregate counts for the following tumor sites: Bone, Liver, Lymph Node, Prostate and Soft Tissue to be interesting so we further subsetted the data to explore the distribution of tumors. Additionally, we wanted to further assess whether bone, as mentioned earlier, was indeed the site of greatest metastatic disease. 

```{r}
#calculate the frequency of tumors by age group 
freq <- age_tumor %>% group_by(agegrp,Tumor.Site) %>% summarise(Freq=n())
freq <- rename(freq, "Count of Tumors"=Freq) #rename the frequency column 
freq_gt1 <-freq[which(freq$`Count of Tumors`>1),] #subset the data to include the five columns of interest 
library(ggplot2) #load in ggplot to create graphical summary 
library(viridis) #load in viridis for pretty color palette 
#create a bar chart of the data 
ggplot(freq_gt1, aes(agegrp,`Count of Tumors`)) +
  labs(x="Age Grouped by BJC paper levels", y="Frequency of Tumors") +
  geom_bar(aes(fill=`Tumor.Site`),stat="identity",position="dodge") +
  scale_fill_viridis_d(option  = "mako") 
```

Curiously enough, bone was not the most common site of metastatic disease among this prospective prostate cohort. In fact, the lymph nodes were shown to have a greater quantity of tumors amounting to 61 tumors across the entire cohort as opposed to 43 among bone. Here from this graph we can also see that the "60-69" and "70-79" groups experience the most tumors proportionally, as would  be expected given that as men age the likelihood of developing prostate cancer increases. If time permitted we would have further calculated the incident rate ratios per age group for the five tumor sites where frequency was greater than 1 to understand the proportional burden of disease experienced by each age group. 

Overall, this aim confirms that as men age incidence of prostate cancer increases with metastasis of disease occurring more frequently in the lymph nodes and bone. This data was limited given that the population was a bit skewed toward the older age groups (60-79) with very few cases representing those older than 80. More information as to prognosis and vital status beyond age 80 would be beneficial in determining whether the assumption of age and incidence is valid. 


##Extract the Data from National Cancer Institute 
We downloaded the second cancer dataset from the National Cancer Institute GDC Data Portal for all the clinical cases including all the columns. 

The link can be found [here.](https://portal.gdc.cancer.gov/exploration?cases_size=100&filters=%7B%22op%22%3A%22and%22%2C%22content%22%3A%5B%7B%22op%22%3A%22in%22%2C%22content%22%3A%7B%22field%22%3A%22cases.primary_site%22%2C%22value%22%3A%5B%22prostate%20gland%22%5D%7D%7D%2C%7B%22content%22%3A%7B%22field%22%3A%22cases.project.project_id%22%2C%22value%22%3A%5B%22TCGA-PRAD%22%5D%7D%2C%22op%22%3A%22in%22%7D%2C%7B%22content%22%3A%7B%22field%22%3A%22genes.is_cancer_gene_census%22%2C%22value%22%3A%5B%22true%22%5D%7D%2C%22op%22%3A%22in%22%7D%5D%7D&searchTableTab=cases)

##Excel Cleaning for Second Data Set 
The data was downloaded as a TSV file so it needed to be imported into Excel to change the format to a CSV or .XLSX file. The raw data contained  960 rows and 156 columns. Upon further investigation it was evident that several of the columns were empty or had values missing. In order to determine the columns with no data present along with those which had missing values we used the Filter tool in Excel. Missing values were those marked as '-- in the cells. We developed a key for the missing values called "Key for Missing Values Columns". 

Based on the assessment where we developed the key of the missing values we determined that the raw data contained 112 empty columns for which no data was available and 33 columns with missing values. The 112 empty columns were then removed from the analysis by simple deletion in Excel and a new sheet was created, "Sample". This new sheet "Sample" contained 44 columns. 

Further investigation of the new "Sample" data indicated that some rows were shifted over particularly rows 141, 142, 206, 207, 284, 285, 494 and 495. We pinpointed the origin of the shift for the aforementioned rows to column Y. In order to account for this shift we copied the data from the "Sample" worksheet to a new sheet called "Sample Revised". Here we inserted a cell in column Y for each of the aforementioned rows to shift the values of the cells to the the right so that the cells would be associated with the correct column. 

Similarly the "treatment_type" column values were shifted to the right but also broken up into values of "Radiation Therapy/Pharmaceutical Therapy" and "NOS". To fix this issue we concatenated the values for the therapy and NOS in a new column, column AS. We then named this column "treatment_type".

After concatenating the values we copied the data from "Sample Revised" to a new worksheet called "cleaned_sample". In this dataset we noticed that after shifting the rows five columns were shown to be empty. Thus we removed the five columns: secondary_gleason_grade, synchronous_malignancy, tumor_grade, year_of_diagnosis and treatment_or_therapy.  We exported this data as a CSV file to then continue our analysis of the missing values in R.  

##Dealing with Missing Values 
As mentioned previously, this GDC portal "uncleaned" dataset contained 33 columns with missing values. To be able to adequately analyze the data we need to deal with these missing value strings. First we began by exporting our cleaned data sample from Excel into R. 

```{r}
sample <- read.csv("cleanedsample.csv")
```

Next, we recoded the string values representative of missing values to NAs using the direct substituion with Base R. 

```{r}
temp.NA <- sample
temp.NA[temp.NA == "'--"] <- NA
temp.NA
```
A cursory scan revealed that the strings were indeed changed to NA values within the dataset. 

```{r}
#changing '-- Null values to NA 
temp.NA1 <- subset(temp.NA, temp.NA$classification_of_tumor != "NA" & temp.NA$last_known_disease_status !="NA" & temp.NA$tumor_largest_dimension_diameter!="NA")
temp.NA1

```
```{r}
#explored the relationship between the missing variables sorted by Case ID 
library(VIM)
matrixplot(temp.NA1,sortby=1)

```

```{r}
#created a margin plot to observe how the missing values in days to last follow up comapre with each case 
marginplot(temp.NA1[,c(1,20)])

```

Here we removed the aforementioned columns that split the therapy and NOS which corresponded to columns AK-AN




```{r}
#install.packages("sqldf")
library(sqldf)
new_sample <- read.csv("removed_dups.csv")
new_sample <- subset(new_sample, select=-c(36:39))
dim(new_sample)
GDC <- read.csv("Dataset2DD.csv")
GDC <- rename(GDC, case_id=Case.UUID)
dim(GDC)

dataset_1 <- remove_nonuniques
dataset_1 <- rename(dataset_1, ID=Patient.ID)
dataset_1 <- rename(dataset_1,Age=Diagnosis.Age)
dataset_1sql <- sqldf("SELECT ID, case when Age >=40 and Age <= 59 then '40-59' when Age >59 and Age <= 69 then '60-69'when Age >69 and Age <= 79 then '70-79' else '80+' end  FROM dataset_1 GROUP BY ID,
                      case when Age >=40 and Age <= 59 then '40-59' when 
                      Age >59 and Age <= 69 then '60-69'when
                      Age >69 and Age <= 79 then '70-79' 
                      else '80+' end")
names(dataset_1sql)[2]<-"agegrp"
join_data1_sql <- sqldf("SELECT * FROM dataset_1 JOIN dataset_1sql on dataset_1.ID=dataset_1sql.ID")

#age check 
age_ds1 <- select(join_data1_sql,"Age","agegrp")

joined_data <- sqldf("SELECT * FROM new_sample JOIN GDC on new_sample.case_id=GDC.case_id")
view(joined_data)
dim(joined_data)
```

extracting numeric digits as text mining 

```{r}
z <- substr(joined_data$Age.at.diagnosis, 1,2)
z
joined_data$Age <- cbind(z)
joined_data$Age <- as.numeric(joined_data$Age)
joined_data[36] <- NULL
view(joined_data)
joined_data <- rename(joined_data, ID=case_id)
dim(joined_data)

```
```{r}
repeat_cols <- find_nonuniq(joined_data)
remove_dups <- joined_data[,-repeat_cols]
dim(remove_dups)
```
```{r}
carbon.copy <- function(df){
  df
  for(i in 1:ncol(df)){
    for(j in 1:nrow(df[i])){
      if(!(nrow(unique(df[i])) > 1)){
        df = df[-c(i)]
      }
    }
  }
  return(df)
}

remove_dup_upd <- carbon.copy(remove_dups)

```




```{r}
## removing some duplicated manually 
remove_dups[52]<-NULL #removes the ethnicity duplicate 
remove_dups[52] <- NULL #removes the age duplicate 
dim(remove_dups)
```

```{r}
dataset_2 <- remove_dups
dataset_2sql <- sqldf("SELECT ID, case when Age >=40 and Age <= 59 then '40-59' when Age >59 and Age <= 69 then '60-69'when Age >69 and Age <= 79 then '70-79' else '80+' end  FROM remove_dups GROUP BY ID,
                      case when Age >=40 and Age <= 59 then '40-59' when 
                      Age >59 and Age <= 69 then '60-69'when
                      Age >69 and Age <= 79 then '70-79' 
                      else '80+' end")
names(dataset_2sql)[2]<-"agegrp"
dataset_2$agegrp <- cbind(dataset_2sql[2])
view(dataset_2)

```


```{r}
ggplot(data.frame(dataset_2), aes(x=Age)) +
  geom_histogram(colour="black",fill="lightskyblue") +theme_classic()
  
summary(dataset_2$Age)
  
ggplot(data.frame(dataset_1), aes(x=Age)) +
  geom_histogram(colour="black",fill="lightsalmon") + theme_classic()
summary(dataset_1$Age)

```


```{r}
join_data1_sql[12] <- NULL
```


```{r}
freq_1treat <-join_data1_sql %>% group_by(agegrp,Prior.Treatment) %>% summarise(Freq=n())
dat2 <- dataset_2
dat2[dat2 == "'--"] <- NA
dat2 <- dat2[!is.na(dat2$prior_treatment),]
freq_2treat <- dat2 %>% group_by(agegrp,prior_treatment) %>% summarise(Freq=n())


ggplot(data.frame(freq_1treat), aes(x=agegrp, y=Freq)) +
  labs(x="Age Grouped by BJC paper levels", y="Frequency of Prior Treatment", ) +
  geom_bar(aes(fill=Prior.Treatment),stat="identity",position="dodge") +
  guides(fill=guide_legend(title="Prior Treatment"))+
  scale_fill_brewer(palette="Oranges") +
  theme_classic()

x <- c("40-59","40-59","60-69","60-69","70-79","70-79","80+")
freq_2treat$AGEgrp <- cbind(x)
```


```{r}
colors <-c("#014f86","#bde0fe")
ggplot(data.frame(freq_2treat), aes(x=AGEgrp, y=Freq)) +
  labs(x="Age Grouped by BJC paper levels", y="Frequency of Prior Treatment", ) +
  geom_bar(aes(fill=prior_treatment),stat="identity",position="dodge") +
  guides(fill=guide_legend(title="Prior Treatment"))+
  scale_fill_manual(values=colors)+
  theme_classic()


```







